Exploring Local Explanations of Non-linear Models Using Animated Linear Projections

### 2022-01-06 _paper.tex
typo
done with SHAP and local explanations

\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}

@@

  article]{article}

@@

\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex

@@

  pdftitle={Scrutinizing the linear variable importance of local explanations of non-linear models with animated linear projections},
  pdfkeywords={R, XAI, explanable artificial intelligence, black-box models, local explanations, grand tour, manual tour, radial tour},

@@

  linkcolor=red,
  filecolor=Maroon,
  citecolor=Blue,
  urlcolor=blue,

@@

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

@@

@@

\ifluatex

@@

  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}

\title{Scrutinizing the linear variable importance of local explanations of non-linear models with animated linear projections}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle
\begin{abstract}
Artificial Intelligence (AI) has seen a revitalization in recent years from the use of increasingly hard-to-interpret black-box models. In such models, increased predictive power comes at the cost of the opacity of factor analysis and variable interpretability. The loss of interpretability has led to the field of eXplainable AI (XAI). XAI attempts to shed light on these models by providing means for interpreting them. A \emph{local explanation} of a non-linear model gives a point-estimate of linear variable importance in the vicinity of one observation. After creating a model and exacting explanations for each observation, we explore the original variables, the explanation's attribution, and model information side-by-side with interactively display. After identifying an observation of interest, its local explanation is used as a 1D projection basis. We then manipulate the magnitude of a variable's contribution with a \emph{tour} technique. The tour animates many projections over small changes in the projection basis as it changes the contribution of one variable. Doing so allows an analyst to visually explore the data through the lens of this local explanation and test the variable support where the explanation seems valid. The interactive application and code facilitating this methodology is available are the \textbf{R} package \textbf{cheem}, available on CRAN.
\end{abstract}

@@

There are different reasons and emphases to fit a model. Breiman and Shmueli (Breiman 2001; Shmueli 2010) introduce the idea of distinguishing modeling based on its purpose; \emph{explanatory} modeling is done for some inferential purpose, while \emph{predictive} modeling focuses more on the predictions of out-of-sample observations. The intended use has important implications for model selection and development. In explanatory modeling, interpretability is vital for drawing inferential conclusions. While predictive modeling may opt for more accurate non-linear models. The use of black-box models is becoming increasingly common, but not without their share of controversy (O'neil 2016; Kodiyan 2019). However, the loss of interpretation should not be taken lightly.

Interpretability is vital for exploring and protecting against potential biases in any model. For instance, models regularly pick up on biases in the training data that have observed influence on the response variable, which is then built into the model. Such biases include sex (Dastin 2018; Duffy 2019), race (Larson et al. 2016), and age (Díaz et al. 2018). Variable-level interpretability of models is essential in the evaluation models for such biases.

Another concern is that of data drift. Data drift is an unexpected shift in support of the variables. Non-linear models are regularly more sensitive to these changes. The interpretability of the model means more transparency to see where models' predictions may be plausible or completely unreliable.

Explanabile Artificial Intelligence (XAI) is a recent branch of research that tries to increase the interpretability of black-box models. One way to do so is to use \emph{local explanations}, which take a variable attribution approach to bring transparency to a model. Local explanations attempt to approximate linear variable importance at the location of one observation. Even with a linear space or local approximation, the number of terms is usually sizable, thus challenging to visualize comprehensively.

@@

There are various types of tours distinguished by the generation of projection bases. In a \emph{manual} tour (Cook and Buja 1997; Spyrison and Cook 2020), the path is defined by changing the contribution of a selected variable. Applying tours to models has been done in a couple of contexts. Specifically for exploring various statistical model fits and classification boundaries (Wickham, Cook, and Hofmann 2015), and using tree- and forest-based approaches as a projection pursuit index to generate a tour basis paths (Y. D. Lee et al. 2013; Silva, Cook, and Lee 2021).

The proposed approach uses the radial, manual tour to scrutinize a local explanation. After identifying an observation of interest, its explanation can be evaluated by testing the support of the structure identified by changing variable contribution with the radial tour. We provide a free and open-source R package \textbf{cheem} with an interactive application to facilitate analysis. We supply toy and modern datasets case studies for classification and regression tasks.

The change in the projection basis might feel similar to counterfactual, what-if analysis, such as \emph{ceteris paribus} (Biecek 2020). This phrase, Latin for ``other things held constant'' or ``all else unchanged,'' shows how an observation's prediction would change from a marginal change in one explanatory variable given that other variables are held constant. It ignores correlations of the variables and imagines a case that was not observed. In contrast, our approach is a geometric explanation of the factual; it varies contributions of the variables by rotating the basis, a reorientation of the data object. Another difference is that the basis must remain orthonormal. That is to say, when the contribution of one variable decreases, the contributions of others necessarily increase such that there is a complete component in that direction.

The remainder of this paper is organized as follows. The following section, \protect\hyperlink{sec:explanations}{Local explanation statistics}, covers the background of the local explanation, SHAP, and the traditional visuals produced from it. \protect\hyperlink{sec:tour}{Tours and the radial tour} digs deeper into these animations of continuous linear projections. The section \protect\hyperlink{sec:applicationdesign}{Application Design} discusses the visual layouts, how they facilitate analysis, data preprocessing, and package infrastructure. The section \protect\hyperlink{sec:casestudies}{Case Studies} illustrates several applications of this method. We conclude with a \protect\hyperlink{sec:cheemdiscussion}{Discussion} of the insights we draw from classification and regression tasks.

\hypertarget{sec:explanations}{%
\section{Local explanation statistics}\label{sec:explanations}}

Consider a highly non-linear model. At face value, it is hard to say which variable(s) are sensitive to crossing a classification boundary or identify which variables caused an observation to have a relatively extreme residual. Local explanations shed light on these cases by approximating linear variable importance in the vicinity of one observation.

Figure six of Arrieta et al. (2020) gives comprehensive summarization of the taxonomy and literature of explanation techniques. The figure includes a large number of model-specific explanations such as deepLIFT, (Shrikumar et al. 2016; Shrikumar, Greenside, and Kundaje 2017) a popular recursive method for estimating importance in neural networks. There are fewer model-agnostic explanations, of which LIME, (Ribeiro, Singh, and Guestrin 2016) SHAP, (S. Lundberg and Lee 2017) and their variants are popular.

These instance-level explanations are used in various ways depending on the data. In images, saliency maps overlay or offset a heatmap indicating necessary pixels (Simonyan, Vedaldi, and Zisserman 2014). For instance, snow may be highlighted when distinguishing if a picture contains a wolf or husky. In text analysis, word-level contextual sentiment analysis can be used to highlight the sentiment and magnitude of influential words (Vanni et al. 2018). In the case of numeric regression, they are used to explain variable additive contributions from the model intercept to the observation's prediction (Ribeiro, Singh, and Guestrin 2016).

\hypertarget{shap-and-tree-shap}{%
\subsection{SHAP and tree SHAP}\label{shap-and-tree-shap}}

SHaply Additive exPlanations (SHAP) approximates the variable importance in the vicinity of one observation by taking the median importance of a subset of permutations in the explanatory variables. This idea stems from the field of game theory, where Shapley (1953) devised a method to evaluate an individual's contribution to cooperative games by permuting the players that contribute to the score.

To illustrate SHAP and its original use, explaining the difference between the intercept and an observation's prediction, we use soccer data from FIFA 2020 season (Leone 2020). We have 5000 observations of nine skill measures (after aggregating highly correlated variables). A random forest model is fit to regress the log wages, in 2020 Euros, from the skill measures. We then extract the SHAP values of a star offensive player (L. Messi) and defensive player (V. van Dijk). We expect to see a difference in the attribution of the variable importance across the two positions of the players.

@@

\caption{Illustration of the distribution of SHAP attributions and a break-down plot. From FIFA 2020 data, a random forest model regresses wages from nine skill attributes for a star offensive and defensive player. The players have very different salaries, but a) shows the distributions of 25 permutations in the explanatory variables. The medians of these distributions are the final SHAP values. The variable importance differs across the exogenous information of player position. These explanations make sense; the variable importances seem realistic given the player's positions. b) Break-down plots of the observations using their explanations to additively cover the difference between the model intercept and the observation predictions.}\label{fig:shapdistrbd}
\end{figure}

Figure \ref{fig:shapdistrbd} shows the SHAP values of these players. Panel a) shows these players receive a sizable difference in wages. Panel b) shows the underlying distribution of the SHAP attributions while permuting the explanatory variables, with the medians being the SHAP values. In the light of the player position, the difference in the variable importance makes sense; offense and movement skills are more important for the offensive player, while defensive and power skills are more informative to the model for explaining the prediction of the defensive player. We would likewise expect the profile of variable importance to be unique for star players of other positions, such as goalkeepers or middle fielders. Panel b) shows a simplified break-down plot (Gosiewska and Biecek 2019), where a local explanation is used to additively explain the difference from the intercept to the observation's prediction. Such additive approaches will show asymmetry in the variable ordering, so we opt to fix the order to panel a) by decreasing the sum of the SHAP values.

In summary, this figure highlights how local explanations bring interpretability to a model, at least in the vicinity of their observations. In this instance, two players with different positions receive different profiles of variable importance to explain the prediction of their wages. In application, we apply \emph{tree SHAP}, a variant of SHAP enjoys a lower computational complexity (S. M. Lundberg, Erion, and Lee 2018). Tree SHAP is only compatible with tree-based models; we illustrate random forests. The following section will use normalized explanations as the starting projection basis to further scrutinize the explanation.

@@

\hypertarget{manual-tours-and-its-radial-case}{%
\subsection{Manual tours and its radial case}\label{manual-tours-and-its-radial-case}}

@@

\hypertarget{sec:applicationdesign}{%
\section{cheem viewer}\label{sec:applicationdesign}}

@@

\begin{itemize}
    \item \textbf{Data:} a complete numerical matrix; explanatory and response variable. An optional categorical variable can be mapped to the color and shape of observations. Explanatory variables are scaled in visualization after modeling or creating local explanations. 
    \item \textbf{Model:} any model can be used with this method. Currently, we apply random forest models via the package \textbf{randomForest} [Liaw and Wiener (2002)] for compatibility with the local explanation, which requires tree-based models.
    \item \textbf{Local explanation:} any model-compatible linear explanation could be used. We apply tree SHAP, a more computationally efficient variant of SHAP, compatible with tree-based models. Tree SHAP was calculated with the package \textbf{treeshap} [Kominsarczyk et al. (2021), hosted on GitHub only]. The global view shows all observations in attribution space, requiring the variable importance from \emph{all} observations rather than just one.
\end{itemize}

@@

@@

\caption{Species classification of Palmer penguin data. We select a chinstrap penguin that is mislabeled as an adelie. By varying the contribution to bill length, we observe the explanation does not hold when bill length has a significant contribution. The .mp4 animation of this tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_penguins.mp4}{github.com/nspyrison/cheem\_paper/blob/main/figures/case\_penguins.mp4}}\label{fig:casepenguins}
\end{figure}

In figure \ref{fig:casepenguins}, a misclassified point is contrasted with a correctly classified point of its observed class nearby in data-space. The attribution space from the tree SHAP local explanations is a more separable space, where the comparison is squarely in the middle of the orange distribution. The primary observation is between the predicted and observed clusters, a sign of uncertainty in the prediction. The tour varies the contribution of bill length (b\_l) as this variable differs most from the contribution of the comparison observation. Downplaying the contribution of bill length is crucial to the linear explanation of this observation being misclassified.

@@

@@

\caption{Chocolates data type classification (milk or dark). We select a chocolate labeled as dark though a random forest model predicts it to be milk chocolate in light of the values on the nutritional label. We vary the contribution of calories from fat. Animated tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_chocolates.mp4}{github.com/nspyrison/cheem\_paper/blob/main/figures/case\_chocolates.mp4}.}\label{fig:casechocolates}
\end{figure}

Figure \ref{fig:casechocolates} similarly shows that attribution-space is more separable than data-space. Interestingly, the class imbalance that we suspected was not observed; there are only six chocolates labeled as dark and predicted as milk, while eight of the inverse case. Calories from fat is the variable with the largest difference in treeshap attribution between these points. In this case, it feels strange to call the selected observation a misclassification of the model. There are plausible reasons that a manufacturer has incentives to cut corners and label their products different than what they are. This feels more like a measurement theory-related problem. In this case, the candy is being sold as dark chocolate, while nutritional value more closely resembles milk chocolates.

@@

The 2020 season FIFA data (Leone 2020; Biecek 2018) contains many skill measurements of soccer/football players and wage information. After aggregation of the skill measurements, we regress the log wages {[}2020 euros{]} given just the skill aggregates. The model was fit from 5000 observations of the nine skill aggregates before being thinned to 500 players to mitigate occlusion and render time. We compare a leading offensive fielder (L. Messi) with that of a top defensive fielder (V. van Dijk), the same observations were used in figure \ref{fig:shapdistrbd}.

@@

\caption{FIFA 2020, regressing log wages {[}2020 Euros{]} from aggregations of skill measurements. The primary observation is a star offensive player (L. Messi) compared with a top defensive player (V. van Dijk). The animate radial tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_fifa.mp4}{github.com/nspyrison/cheem\_paper/blob/main/figures/case\_fifa.mp4}}\label{fig:casefifa}
\end{figure}

With figure \ref{fig:casefifa}, we will test the premise of the local explanation. If we remove reaction and movement skills from the basis, offense skills are almost singularly important for explaining the offensive player. We vary the contribution of offensive skills. Offensive skills are removed in the tour (panel b, frame 3), and Messi is no longer separated from the group. We also notice that accuracy has rotated into the frame, maintaining some separability.

@@

@@

\caption{Ames housing 2018 regressing log sales price {[}2018 USD{]}. Because the SHAP values are relatively well distributed across the variables, there is a lot of redundant information to explain the gap between select observations. This case may not be ideal for the cheem methodology, though we did find a near singular frame in the manual tour not, something we came looking for, but certainly, an exciting feature to find. The corresponding animation is at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_ames2018.mp4}{github.com/nspyrison/cheem\_paper/blob/main/figures/case\_ames2018.mp4}}\label{fig:caseames}
\end{figure}

Figure \ref{fig:caseames} shows the global view and extrema of the tour. The horizontal distance in the tour did not show a significant disparity between our selected points. This is not particularly surprising as most variables have a sizable contribution. Rotating any one variable out of the frame will rotate other vital variables into the frame, preserving most of the distance from intercept to prediction. However, the tour has revealed an interesting feature worth discussing. Notice that the observations pivot about the origin, the basis roughly halfway between bases in frames one and two of panel b) the data is near a singular profile. This means that there is a basis orthogonal to this point that describes sizable variation. Knowing these singular bases can point toward others with meaningful data variation.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-arrieta_explainable_2020}{}%
Arrieta, Alejandro Barredo, Natalia Díaz-Rodríguez, Javier Del Ser, Adrien Bennetot, Siham Tabik, Alberto Barbado, Salvador García, Sergio Gil-López, Daniel Molina, and Richard Benjamins. 2020. {``Explainable {Artificial} {Intelligence} ({XAI}): {Concepts}, Taxonomies, Opportunities and Challenges Toward Responsible {AI}.''} \emph{Information Fusion} 58: 82--115.

\leavevmode\hypertarget{ref-asimov_grand_1985}{}%
Asimov, Daniel. 1985. {``The {Grand} {Tour}: A {Tool} for {Viewing} {Multidimensional} {Data}.''} \emph{SIAM Journal on Scientific and Statistical Computing} 6 (1): 128--43. https://doi.org/\url{https://doi.org/10.1137/0906011}.

\leavevmode\hypertarget{ref-biecek_dalex_2018}{}%
Biecek, Przemyslaw. 2018. {``{DALEX}: Explainers for Complex Predictive Models in {R}.''} \emph{The Journal of Machine Learning Research} 19 (1): 3245--49.

\leavevmode\hypertarget{ref-biecek_ceterisparibus_2020}{}%
---------. 2020. \emph{{ceterisParibus}: {Ceteris} {Paribus} {Profiles}}. \url{https://CRAN.R-project.org/package=ceterisParibus}.

\leavevmode\hypertarget{ref-biecek_explanatory_2021}{}%
Biecek, Przemyslaw, and Tomasz Burzykowski. 2021. \emph{Explanatory {Model} {Analysis}: {Explore}, {Explain}, and {Examine} {Predictive} {Models}}. CRC Press.

\leavevmode\hypertarget{ref-breiman_statistical_2001}{}%
Breiman, Leo. 2001. {``Statistical Modeling: {The} Two Cultures (with Comments and a Rejoinder by the Author).''} \emph{Statistical Science} 16 (3): 199--231.

>>>>

\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}

@@

]{article}

@@

\usepackage{iftex}
\ifPDFTeX

@@

  pdftitle={Exploring Local Explanations of Non-linear Models Using Animated Linear Projections},
  pdfkeywords={R, XAI, explanable artificial intelligence, black-box
models, local explanations, grand tour, manual tour, radial tour},

@@

  linkcolor={red},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},

@@

@@

\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \setlength{\parskip}{#2\cslentryspacingunit}

@@

\ifLuaTeX

@@

\title{Exploring Local Explanations of Non-linear Models Using Animated
Linear Projections}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle
\begin{abstract}
Artificial Intelligence (AI) has seen a revitalization in recent years
with the prevalence of increasingly hard-to-interpret black-box models.
The increased predictive power comes at the cost of interpretability,
which has led to the emergence of eXplainable AI (XAI). XAI attempts to
shed light on how models are using predictors, to arrive at a
prediction, with a point estimate of the linear variable importance in
the vicinity of each observation. These are simply linear projections,
and as such can be further explored interactively to better understand
the interaction between variables used to make predictions, across the
predictive model surface. Here we describe interactive linear
interpolation used for exploration at any observation, and illustrate
with examples with categorical (penguin species, chocolate types) and
quantitative (footbal salaries, house prices) response variables. The
methods are implemented in the \textbf{R} package \textbf{cheem},
available on CRAN.
\end{abstract}

@@

There are different reasons and emphases to fit a model. Breiman (2001),
reiterated by Shmueli (2010), taxonomizes modeling based on its purpose;
\emph{explanatory} modeling is done for some inferential purpose, while
\emph{predictive} modeling focuses more on the predictions of
out-of-sample observations. The intended use has important implications
for model selection and development. In explanatory modeling,
interpretability is vital for drawing inferential conclusions. While
predictive modeling may opt for more accurate non-linear models. The use
of black-box models is becoming increasingly common, but not without
their share of controversy (O'Neil 2016; Kodiyan 2019). However, the
loss of interpretation presents a challenge.

Interpretability is vital for exploring and protecting against potential
biases (e.g.~sex (Dastin 2018; Duffy 2019), race (Larson et al. 2016),
and age (Díaz et al. 2018)) in any model. For instance, models regularly
pick up on biases in the training data that have observed influence on
the response (output) variable, which is then built into the model.
Variable-level (feature-level) interpretability of models is essential
in the evaluation models for such biases. It is also generally important
for many problems, where it is not enough to accurately predict
accurately but one must be able to explain which predictors are most
responsible in generating a response value.

Another concern is that of data drift, which is a shift in support or
domain of the explanatory variables (feature or predictors). Non-linear
models are typically more sensitive, and do not extrapolate well outside
of the training data domain. Better interpretability of the model means
that there is more transparency where models' predictions may be
plausible or completely unreliable.

Explainable Artificial Intelligence (XAI) is an emerging field of
research that tries to increase the interpretability of black-box
models. A common approach is to use \emph{local explanations}, which
attempt to approximate linear variable importance at the location of
each observation (instance), or the predictions at a specific point in
the data domain. Because these are point specific, a challenge is
visualizing them to more comprehensively understand a model.

XXX DC Note that we need to make sure that the paper is readable for a
machine learner, so a translation of terms is important. I've done this
by parenthesizing the ML term in this introduction, which should be
enough.

@@

There are various types of tours distinguished by the generation of
projection bases. In a \emph{manual} tour (Cook and Buja 1997; Spyrison
and Cook 2020), the path is defined by changing the contribution of a
selected variable. Applying tours to models has been done in a couple of
contexts. Specifically for exploring various statistical model fits and
classification boundaries (Wickham, Cook, and Hofmann 2015), and using
tree- and forest-based approaches as a projection pursuit index to
generate a tour basis paths (Y. D. Lee et al. 2013; da Silva, Cook, and
Lee 2021).

In our proposed approach we use the radial manual tour to scrutinize a
local explanation. Additional interactivity allows the user to identify
an observation of interest, then explore its local explanation by
changing variable contribution with the radial tour. The methods are
implemented in R package \textbf{cheem}. Example datasets are provided
to illustrate usage for classification and regression tasks.

Using a radial tour can be considered similar to counterfactual, what-if
analysis, such as \emph{ceteris paribus} (Biecek 2020). This phrase,
Latin for ``other things held constant'' or ``all else unchanged'',
shows how an observation's prediction would change from a marginal
change in one explanatory variable given that other variables are held
constant. It ignores correlations of the variables and imagines a case
that was not observed. In contrast, our approach is a geometric
explanation of the factual; it varies contributions of the variables by
rotating the basis, a reorientation of the data object. A constraint in
our approach is that the basis must remain orthonormal. That means, when
the contribution of one variable decreases, the contributions of others
necessarily increase such that there is a complete component in that
direction. This also ensures that what is seen is strictly a
low-dimensional projection from high-dimensions, and is thus an
interpretable visualisation.

The remainder of this paper is organized as follows. The following
section, \protect\hyperlink{sec:explanations}{Local explanations} covers
the background of the local explanation, and the traditional visuals
produced. \protect\hyperlink{sec:tour}{Tours and the radial tour}
explains the animations of continuous linear projections. The section
\protect\hyperlink{sec:applicationdesign}{Application Design} discusses
the visual layout in the interactive interface, how they facilitate
analysis, data preprocessing, and package infrastructure. The section
\protect\hyperlink{sec:casestudies}{Case Studies} illustrates the
application to supervised learning with categorical and quantitative
response variables. We conclude with a
\protect\hyperlink{sec:cheemdiscussion}{Discussion} of the insights
gained and directions that might be explore in the future. (XXX DC Can
you change the references to number of the section not title?)

\hypertarget{sec:explanations}{%
\section{Local explanations}\label{sec:explanations}}

Consider a highly non-linear model. It can be hard to determine whether
small changes in a variable's value will make a class prediction change
group or identify which variables contribute to an extreme residual.
Local explanations shed light on these situations by approximating
linear variable importance in the vicinity of a single observation or
point.

A comprehensive summary of the taxonomy and literature of explanation
techniques is provided in Figure 6 of Arrieta et al. (2020). It includes
a large number of model-specific explanations such as deepLIFT
(Shrikumar et al. 2016; Shrikumar, Greenside, and Kundaje 2017), a
popular recursive method for estimating importance in neural networks.
There are fewer model-agnostic explanations, of which LIME (Ribeiro,
Singh, and Guestrin 2016) SHAP (S. Lundberg and Lee 2017), and their
variants are popular.

These instance-level explanations are used in various ways depending on
the data. In image classification, where pixels would correspond to
predictors, saliency maps overlay or offset a heatmap indicating
important pixels (Simonyan, Vedaldi, and Zisserman 2014). For instance,
pixels corresponding to snow may be highlighted when distinguishing if a
picture contains a wolf or husky. In text analysis, word-level
contextual sentiment analysis can be used to highlight the sentiment and
magnitude of influential words (Vanni et al. 2018). In the case of
numeric regression, they are used to explain variable additive
contributions from the model intercept to the observation's prediction
(Ribeiro, Singh, and Guestrin 2016).

SHaply Additive exPlanations (SHAP) approximates the variable importance
in the vicinity of one observation conceptually by examining the effect
of other variables on the contribution of the variable of interest on
predicting the response. This explanations almost all point to Shapley
(1953)'s method to evaluate an individual's contribution to cooperative
games by permuting the players that contribute to the score. Strumbelj
and Kononenko (2010) introduced the use of SHAP for local explanations
in ML models. However, it is also related to partial dependence plots
(XXX Needs reference), used to explain the effect of a variable by
predicting the response for a range of values on this variable, after
fixing the value of all other variables to their mean. It could also be
considered to be similar to examining the coefficients from all subsets
regression, as described in Wickham, Cook, and Hofmann (2015), which
helps to understand the relative importance of each variable in the
context of all other candidate variables.

For our application, we use \emph{tree SHAP}, a variant of SHAP enjoys a
lower computational complexity (S. M. Lundberg, Erion, and Lee 2018).
Tree SHAP is only compatible with tree-based models; we illustrate
random forests. The following section will use normalized explanations
as the starting projection basis to further scrutinize the explanation.

XXX DC does this example come from/motivated by
\url{https://ema.drwhy.ai/UseCaseFIFA.html} ?? it needs to be cited
here. To understand the use SHAP we use soccer data from FIFA 2020
season (Leone 2020). There are 5000 observations of 9 skill measures
(after aggregating highly correlated variables). A random forest model
is fit with log wages, in 2020 Euros, as the response, using the skill
measures as predictors. We then extract the SHAP values of a star
offensive player (L. Messi) and defensive player (V. van Dijk). The
results are displayed in Figure @ref(fig:shapdistrbd). We expect to see
a difference in the attribution of the variable importance across the
two positions of the players, which would be interpreted as how the
player's salary depends on this combination of skill sets. Plot (b) is a
modified breakdown plot (Gosiewska and Biecek 2019) where the order of
variables is fixed, so the two observations can be more easily compared.

@@

\caption{Illustration of the use of SHAP values for a random forest model for salaries for FIFA 2020 players based on nine predictors corresponding to different skills. Two observations (a star offensive player, Messi and a defensive player, van Dijk) are compared. Plot (a) shows the distribution of contributions for each variable across 25 permutations of predictors, with the median displayed as a dot, for both players. Plot b) is the break-down plot showing the SHAP values. It can be learned that reaction is important for explaining both players salaries. Offense and movement are important for Messi but not van Dijk, and conversely defense and power are important for van Dijk but not Messi.}\label{fig:shapdistrbd}
\end{figure}

XXX DC Can you remove ``Variable'' from the vertical axes? And remove
reference to ``permutations'' in the figure text. Also, why include
interaction and prediction - suggest removing these because there is no
information. Perhaps switch the order, and put the breakdown plot first?

In summary, these plots highlight how local explanations bring
interpretability to a model, at least in the vicinity of their
observations. In this instance, two players with different positions
receive different profiles of variable importance to explain the
prediction of their wages.

@@

@@

\hypertarget{sec:applicationdesign}{%
\section{The cheem viewer}\label{sec:applicationdesign}}

@@

(ref:citeRf) Liaw and Wiener (2002) (ref:citeTs) Kominsarczyk et al.
(2021)

\begin{itemize}
    \item \textbf{Data:} a complete numerical matrix; explanatory and response variable. An optional categorical variable can be mapped to the color and shape of observations. Explanatory variables are scaled in visualization after modeling or creating local explanations. 
    \item \textbf{Model:} any model can be used with this method. Currently, we apply random forest models via the package \textbf{randomForest} [(ref:citeRf)] for compatibility with the local explanation, which requires tree-based models.
    \item \textbf{Local explanation:} any model-compatible linear explanation could be used. We apply tree SHAP, a more computationally efficient variant of SHAP, compatible with tree-based models. Tree SHAP was calculated with the package \textbf{treeshap} [(ref:citeTs), hosted on GitHub only]. The global view shows all observations in attribution space, requiring the variable importance from \emph{all} observations rather than just one.
\end{itemize}

@@

(ref:casepenguins) Species classification of Palmer penguin data. We
select a chinstrap penguin that is mislabeled as an adelie. By varying
the contribution to bill length, we observe the explanation does not
hold when bill length has a significant contribution. The .mp4 animation
of this tour can be found at
\href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_penguins.mp4}{github.com/nspyrison/cheem\_paper/blob/main/figures/case\_penguins.mp4}

@@

\caption{(ref:casepenguins)}\label{fig:casepenguins}
\end{figure}

In figure @ref(fig:casepenguins), a misclassified point is contrasted
with a correctly classified point of its observed class nearby in
data-space. The attribution space from the tree SHAP local explanations
is a more separable space, where the comparison is squarely in the
middle of the orange distribution. The primary observation is between
the predicted and observed clusters, a sign of uncertainty in the
prediction. The tour varies the contribution of bill length (b\_l) as
this variable differs most from the contribution of the comparison
observation. Downplaying the contribution of bill length is crucial to
the linear explanation of this observation being misclassified.

@@

(ref:casechocolates) Chocolates data type classification (milk or dark).
We select a chocolate labeled as dark though a random forest model
predicts it to be milk chocolate in light of the values on the
nutritional label. We vary the contribution of calories from fat.
Animated tour can be found at
\href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_chocolates.mp4}{github.com/nspyrison/cheem\_paper/blob/main/figures/case\_chocolates.mp4}.

@@

\caption{(ref:casechocolates)}\label{fig:casechocolates}
\end{figure}

Figure @ref(fig:casechocolates) similarly shows that attribution-space
is more separable than data-space. Interestingly, the class imbalance
that we suspected was not observed; there are only six chocolates
labeled as dark and predicted as milk, while eight of the inverse case.
Calories from fat is the variable with the largest difference in
treeshap attribution between these points. In this case, it feels
strange to call the selected observation a misclassification of the
model. There are plausible reasons that a manufacturer has incentives to
cut corners and label their products different than what they are. This
feels more like a measurement theory-related problem. In this case, the
candy is being sold as dark chocolate, while nutritional value more
closely resembles milk chocolates.

@@

The 2020 season FIFA data (Leone 2020; Biecek 2018) contains many skill
measurements of soccer/football players and wage information. After
aggregation of the skill measurements, we regress the log wages {[}2020
euros{]} given just the skill aggregates. The model was fit from 5000
observations of the nine skill aggregates before being thinned to 500
players to mitigate occlusion and render time. We compare a leading
offensive fielder (L. Messi) with that of a top defensive fielder (V.
van Dijk), the same observations were used in figure
@ref(fig:shapdistrbd).

(ref:casefifa) FIFA 2020, regressing log wages {[}2020 Euros{]} from
aggregations of skill measurements. The primary observation is a star
offensive player (L. Messi) compared with a top defensive player (V. van
Dijk). The animate radial tour can be found at
\href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_fifa.mp4}{github.com/nspyrison/cheem\_paper/blob/main/figures/case\_fifa.mp4}

@@

\caption{(ref:casefifa)}\label{fig:casefifa}
\end{figure}

With figure @ref(fig:casefifa), we will test the premise of the local
explanation. If we remove reaction and movement skills from the basis,
offense skills are almost singularly important for explaining the
offensive player. We vary the contribution of offensive skills.
Offensive skills are removed in the tour (panel b, frame 3), and Messi
is no longer separated from the group. We also notice that accuracy has
rotated into the frame, maintaining some separability.

@@

(ref:caseames) Ames housing 2018 regressing log sales price {[}2018
USD{]}. Because the SHAP values are relatively well distributed across
the variables, there is a lot of redundant information to explain the
gap between select observations. This case may not be ideal for the
cheem methodology, though we did find a near singular frame in the
manual tour not, something we came looking for, but certainly, an
exciting feature to find. The corresponding animation is at
\href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_ames2018.mp4}{github.com/nspyrison/cheem\_paper/blob/main/figures/case\_ames2018.mp4}

@@

\caption{(ref:caseames)}\label{fig:caseames}
\end{figure}

Figure @ref(fig:caseames) shows the global view and extrema of the tour.
The horizontal distance in the tour did not show a significant disparity
between our selected points. This is not particularly surprising as most
variables have a sizable contribution. Rotating any one variable out of
the frame will rotate other vital variables into the frame, preserving
most of the distance from intercept to prediction. However, the tour has
revealed an interesting feature worth discussing. Notice that the
observations pivot about the origin, the basis roughly halfway between
bases in frames one and two of panel b) the data is near a singular
profile. This means that there is a basis orthogonal to this point that
describes sizable variation. Knowing these singular bases can point
toward others with meaningful data variation.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-arrieta_explainable_2020}{}}%
Arrieta, Alejandro Barredo, Natalia Díaz-Rodríguez, Javier Del Ser,
Adrien Bennetot, Siham Tabik, Alberto Barbado, Salvador García, Sergio
Gil-López, Daniel Molina, and Richard Benjamins. 2020. {``Explainable
{Artificial} {Intelligence} ({XAI}): {Concepts}, Taxonomies,
Opportunities and Challenges Toward Responsible {AI}.''}
\emph{Information Fusion} 58: 82--115.

\leavevmode\vadjust pre{\hypertarget{ref-asimov_grand_1985}{}}%
Asimov, Daniel. 1985. {``The {Grand} {Tour}: A {Tool} for {Viewing}
{Multidimensional} {Data}.''} \emph{SIAM Journal on Scientific and
Statistical Computing} 6 (1): 128--43.
https://doi.org/\url{https://doi.org/10.1137/0906011}.

\leavevmode\vadjust pre{\hypertarget{ref-biecek_dalex_2018}{}}%
Biecek, Przemyslaw. 2018. {``{DALEX}: Explainers for Complex Predictive
Models in {R}.''} \emph{The Journal of Machine Learning Research} 19
(1): 3245--49.

\leavevmode\vadjust pre{\hypertarget{ref-biecek_ceterisparibus_2020}{}}%
---------. 2020. \emph{{ceterisParibus}: {Ceteris} {Paribus}
{Profiles}}. \url{https://CRAN.R-project.org/package=ceterisParibus}.

\leavevmode\vadjust pre{\hypertarget{ref-biecek_explanatory_2021}{}}%
Biecek, Przemyslaw, and Tomasz Burzykowski. 2021. \emph{Explanatory
{Model} {Analysis}: {Explore}, {Explain}, and {Examine} {Predictive}
{Models}}. CRC Press.

\leavevmode\vadjust pre{\hypertarget{ref-breiman_statistical_2001}{}}%
Breiman, Leo. 2001. {``Statistical Modeling: {The} Two Cultures (with
Comments and a Rejoinder by the Author).''} \emph{Statistical Science}
16 (3): 199--231.
